\topskip0pt
\vspace*{15em}

\begin{center}
    \section*{Abstract}

    Figurative language, such as idioms and metaphors, is a barrier to text accessibility. Easy-to-Read guidelines, which serve readers with intellectual disabilities, low literacy, or limited language proficiency, instruct writers to avoid figurative language but give no method for rewriting it. This thesis studies whether large language models can fill that gap by detecting figurative expressions and rewriting them into literal, meaning-preserving, easier-to-read English, without task-specific training.

    We build an agentic pipeline that decomposes the task into three explicit, inspectable steps (detection, explanation, and transformation) and compare it against a monolithic single-prompt baseline. Both run on open-weights models at two scales, an 8B deliverable and a 70B comparator, and are evaluated on SemEval-2022 Task 2 idioms and the VU Amsterdam Metaphor Corpus metaphors, using automatic detection metrics and a Direct Assessment human-evaluation survey for replacement quality.

    The central finding is that reliable quality comes from explicit task structure rather than from model scale or prompt wording. Scale lifts detection but not human-perceived replacement quality: in this setup, scale did not buy, and on simplicity may have cost, human-perceived quality. Gains from prompt wording do not transfer across model sizes and can reverse in sign. Decomposition yields a small but consistent advantage on meaning preservation, falling just short of statistical significance, with a tied composite score; its clearer benefit is that each failure can be traced to a named step. Across all systems the replacements preserve meaning but do not simplify, which marks the central remaining challenge for Easy-to-Read adaptation, and shows that explicit structure, together with human-grounded evaluation, is the lever that can be relied on in this task.

    The thesis contributes an agentic open-weights pipeline for this task with a monolithic baseline as a designed control, an adapted Direct Assessment protocol for evaluating figurative-to-literal replacement, and a by-product dataset of 108 human-written alternative replacements with rubric ratings for 81 system outputs.

\end{center}


\phantomsection
\addcontentsline{toc}{section}{Abstract}

\newpage