\section{Introduction}
\label{ch:introduction}

% ============================================================================
% === WRITING GUIDE: Chapter 1, Introduction (~4 pages, ~1,800 words) =======
% ============================================================================
% JOB: establish the "missing E2R guideline" motivation, state the central
% thesis, present RQ1-RQ3 with hypotheses (H4 flagged untested up front), and
% list contributions. This chapter carries Mari Carmen's deferred review items
% "storyline framing" and "FACILE positioning" first.
%
% CENTRAL THESIS STATEMENT (recurring thread; open with a version of this):
%   When LLMs adapt figurative language into Easy-to-Read English, reliable
%   quality does not come from where current practice looks for it (larger
%   models, cleverer prompts) but from explicit structure: decomposing the
%   task into inspectable steps and grounding evaluation in human judgement.
%   The lever that pays is structure, in both the system and its measurement.
%
% WRITE THIS CHAPTER AFTER Discussion 8.1 (per WRITING_PLAN.md schedule),
% so the framing cannot overstate the evidence.
% ============================================================================

\subsection{Motivation}
% --- GUIDE (~1.25 pp): figurative language as an accessibility barrier.
% Must land:
%  (1) E2R guidelines explicitly prohibit figurative language, but no system
%      or guideline operationalises HOW to adapt it. Cite the UPM
%      per-guideline-module pattern: \cite{Surez-Figueroa2023} (morphological
%      features), Diab 2024 (dialogues), \cite{Surez-Figueroa2024} (FACILE).
%      This thesis is the figurative-language module of that pattern.
%  (2) The two-literatures gap: figurative-language NLP detects but never
%      adapts for accessibility; E2R automates other guidelines but not this
%      one. Neither community addresses the intersection (developed in Ch 2).
%  (3) Why LLMs now: supervised detection systems exist but replacement has
%      no training data and no automatic metric; zero/one-shot LLMs are the
%      only viable approach for the generative half of the task.

Easy-to-Read English is a way of writing meant for people who find standard
text hard to follow, including people with intellectual disabilities, people
with low literacy, and people reading in a language that is not their first.
The guidelines for this style share one rule: avoid figurative language. Idioms
and metaphors are meant to be removed, because their meaning is not carried by
the words themselves, so a reader who takes them literally is misled and a
reader who cannot recover the intended sense is locked out altogether. In this
setting, figurative language is an accessibility barrier.

The guidelines say to remove such language, but they do not say how to do it,
and they leave the rewriting to a human editor. Recent work at UPM has begun to
automate Easy-to-Read adaptation one guideline at a time, with a dedicated
module for morphological features \cite{Surez-Figueroa2023} and another for
dialogue structure \cite{Diab2024}, part of a research line on guideline-driven
adaptation whose support application, FACILE, is described in
\cite{Surez-Figueroa2024}. Figurative language is
one of the guidelines this programme has not yet covered, and this thesis is
its module: a system that detects an idiom or metaphor and rewrites it into
plain, literal English.

The task sits between two research communities that do not meet. Work on
figurative language in natural-language processing has built strong systems for
finding and labelling idioms and metaphors, but it stops at detection and does
not rewrite the text for a reader. Work on automatic text simplification and
Easy-to-Read adaptation rewrites text for accessibility, but it handles other
guidelines and treats figurative language as out of scope. Neither side
addresses the point where they meet, detecting a figurative expression and
replacing it with an accessible literal form, and Chapter~\ref{ch:background}
develops this gap in full.

Large language models are what make the rewriting half of the task possible
now. Detection alone can be learned from labelled data, and supervised systems
for it already exist, but replacement cannot be learned the same way: there is
no training set of figurative sentences paired with accessible literal
rewrites, and, as this thesis shows, no automatic metric that can score such a
rewrite reliably. A model that can produce a literal restatement from zero or
one example is therefore the only practical approach to the generative half of
the task, and this is the capability that current large language models supply.

\subsection{Problem Statement}
% --- GUIDE (~0.75 pp): state the task (given an English sentence, detect
% idiomatic/metaphorical expressions and produce a literal, meaning-preserving,
% easier-to-read restatement) and then the thesis statement paragraph.
% NOTE: the old bullet "unified treatment of idioms and metaphors" is resolved
% as PARALLEL treatment under a common architecture: four separate LangGraph
% workflows, prompts never shared across phenomena. Say "parallel", not
% "unified" (the implementation deliberately bifurcates).

The task this thesis studies is the following. Given an English sentence,
detect any idiomatic or metaphorical expression in it, and produce a
restatement that keeps the original meaning, removes the figurative wording,
and is easier to read. Detection and rewriting are treated as one connected
job, not two separate ones, because a rewrite is only useful if it lands on
the right expression and a detection is only useful if something then rewrites
it.

Idioms and metaphors are handled in parallel, not as one phenomenon. They
share an architecture but never a prompt: the system runs separate workflows
for each, because an idiom has a fixed conventional meaning while a metaphor
is built from an open mapping between two ideas, and the two call for
different handling. The system is built once and applied twice, with the
difference between the two phenomena kept explicit rather than blended away.

The central claim of this thesis concerns where reliable quality in this task
comes from. Common practice expects two things to drive it: a larger model and
a better-worded prompt. The evidence here points elsewhere. Reliable quality
comes from explicit structure, that is, from breaking the task into
inspectable steps and from grounding the evaluation in human judgement. Model
size and prompt wording are levers whose effect cannot be trusted to hold:
their gains vary across models and can reverse in sign, and the automatic
metrics that might catch the reversal cannot see replacement quality at all.
Explicit task structure is the lever whose effect can be observed, attributed
to a named step, and trusted. Human evaluation is not itself a lever on
quality; it is how that quality is made visible. This is what the thesis means
by structure over scale: not that structure always wins, but that it is the
lever that can be relied on to mean what it says
(Section~\ref{sec:discussion-synthesis}).

\subsection{Research Questions and Hypotheses}
% --- GUIDE (~1 pp): RQ1-RQ3 verbatim from wiki/thesis/research-questions.md:
%  RQ1: To what extent can a zero-/one-shot agentic LLM pipeline accurately
%       detect and interpret idiomatic and metaphorical expressions in
%       context, compared to established supervised benchmarks?
%  RQ2: Does an explicit agentic decomposition (detection -> explanation ->
%       transformation) improve figurative-language handling compared to
%       monolithic prompt-based LLM approaches?
%  RQ3: How effectively can an agentic LLM system replace idiomatic and
%       metaphorical expressions with literal, meaning-preserving
%       alternatives, and what trade-offs arise between semantic fidelity
%       and readability?
% Hypotheses H1-H4. H4 (explanation quality correlates with replacement
% correctness) MUST be stated here and explicitly flagged as untested in this
% thesis, with a pointer to the proposed design in the Discussion (8.5).
% Observability is framed as a design property examined qualitatively in the
% Discussion, NOT as a fourth RQ.

The thesis is organised around three research questions.

\begin{description}
  \item[RQ1 (Detection and interpretation).] To what extent can a
    zero-/one-shot agentic LLM pipeline accurately detect and interpret
    idiomatic and metaphorical expressions in context, compared to established
    supervised benchmarks?
  \item[RQ2 (Agentic decomposition).] Does an explicit agentic decomposition
    (detection, then explanation, then transformation) improve
    figurative-language handling compared to monolithic prompt-based LLM
    approaches?
  \item[RQ3 (Replacement quality).] How effectively can an agentic LLM system
    replace idiomatic and metaphorical expressions with literal,
    meaning-preserving alternatives, and what trade-offs arise between semantic
    fidelity and readability?
\end{description}

Four hypotheses make these questions testable.

\begin{description}
  \item[H1.] A zero-/one-shot agentic LLM pipeline can reach performance
    comparable to supervised systems on idiom detection and metaphor
    identification, despite using no task-specific training.
  \item[H2.] Agentic decomposition improves figurative-language replacement
    quality compared to single-prompt baselines, particularly for metaphors.
  \item[H3.] Idiomatic expressions are replaced with meaning-preserving literal
    alternatives more reliably than conceptual metaphors, because metaphors
    rest on an open mapping between an abstract idea and a concrete one.
  \item[H4.] Explicit intermediate explanations generated by the agent
    correlate positively with the semantic correctness of the final literal
    replacement.
\end{description}

H1 through H3 are tested in the empirical chapters. H4 is not tested in this
thesis. Measuring it requires a separate study that scores explanation quality
and replacement correctness independently and then correlates them, and that
study was out of scope here. H4 is kept on the record because it names the next
natural step, and the design needed to test it is set out as future work in
Section~\ref{sec:discussion-h4}.

A fourth concern, whether the system's steps are observable enough to explain
its own outputs, runs through the work but is not posed as a research question.
It is a property of the design rather than a hypothesis to confirm or reject,
so it is examined qualitatively in the Discussion
(Section~\ref{sec:discussion-observability}) rather than tested.

\subsection{Contributions}
% --- GUIDE (~0.5 pp): four concrete contributions:
%  (1) An agentic open-weights pipeline for figurative-language E2R
%      adaptation (FastAPI + LangGraph + vLLM; 200+ persisted, reproducible
%      runs), with a monolithic baseline as designed control.
%  (2) An adapted Direct Assessment human-evaluation methodology for
%      figurative-language replacement (Ch 4), with documented departures
%      from Graham 2013/2017.
%  (3) Empirical findings on what does and does not transfer: scale lifts
%      detection but not human-perceived replacement quality; prompt gains
%      invert across scales; decomposition buys meaning preservation.
%  (4) A by-product dataset: 108 human-written alternative replacements +
%      rubric ratings for 81 system outputs.

This thesis makes four contributions.

\begin{enumerate}
  \item An agentic open-weights pipeline for figurative-language Easy-to-Read
    adaptation, built on FastAPI, LangGraph, and vLLM, with over two hundred
    persisted and reproducible runs, and a monolithic single-prompt system as a
    designed control.
  \item An adapted Direct Assessment human-evaluation method for
    figurative-language replacement (Chapter~\ref{ch:methodology}), building on
    the established Direct Assessment protocol
    \cite{Graham2013, graham-2015-improving, Graham2017}
    with its departures from that protocol documented, since no automatic
    metric scores this task reliably.
  \item Empirical findings on what does and does not transfer in this task:
    scale lifts detection but not human-perceived replacement quality, prompt
    gains can reverse in sign across model sizes, and decomposition buys a
    consistent directional advantage on meaning preservation.
  \item A by-product dataset of 108 human-written alternative replacements
    together with rubric ratings for 81 system outputs, available for future
    work on this task.
\end{enumerate}

\subsection{Document Structure}
% --- GUIDE (~0.5 pp): one sentence per chapter. Close by framing the three
% empirical chapters as testing three levers: scale (RQ1), structure (RQ2),
% and task difficulty / fidelity-readability trade-offs (RQ3).

The rest of the thesis is organised as follows.
Chapter~\ref{ch:background} reviews the two literatures the task sits between,
figurative-language processing and Easy-to-Read adaptation, and closes by
stating the gap between them.
Chapter~\ref{ch:system-design} describes the agentic pipeline and the
monolithic baseline it is compared against.
Chapter~\ref{ch:methodology} sets out the evaluation method, including the
adapted Direct Assessment survey and the automatic metrics, and explains why
the human survey is required rather than optional.
The three empirical chapters answer one research question each:
Section~\ref{sec:rq1} reports detection and the effect of scale (RQ1),
Section~\ref{sec:rq2} reports the agentic-versus-monolithic comparison (RQ2),
and Section~\ref{sec:rq3} reports replacement quality and the survey results
(RQ3).
Chapter~\ref{ch:discussion} assembles these results into the central claim and
states the limits of the evidence, and
Chapter~\ref{ch:conclusion} closes with the answers to the research questions
and the directions they open.

The three empirical chapters can be read as a test of three levers on quality.
RQ1 tests scale, by comparing an 8B model with a 70B one. RQ2 tests structure,
by comparing a decomposed pipeline with a single prompt. RQ3 tests the task
itself, by measuring how well the rewriting holds meaning while making the text
easier to read, and where those two goals pull against each other. Read
together, they show that the lever that pays is structure.

\newpage
