\section{RQ2: Agentic Decomposition}
\label{sec:rq2}
\textit{Does an explicit agentic decomposition (detection $\to$ explanation $\to$ transformation) improve figurative-language handling compared to monolithic prompt-based LLM approaches?}

% ============================================================================
% === WRITING GUIDE: Chapter 6, RQ2 (~5 pages) ==============================
% ============================================================================
% JOB: deliver the H2 verdict this chapter previously deferred. Automatic
% metrics cannot adjudicate (BLEU and BERTScore point in opposite
% directions); the human evaluation shows a consistent directional advantage
% for the agentic system on meaning preservation, a tied composite, and a
% small monolithic edge on simplicity (n.s.). Concede plainly that
% decomposition buys nothing in detection (Section 6.4).
% MANDATORY HEDGE, verbatim here and wherever H2 is judged (Ch 8, Ch 9):
%   "a consistent directional advantage on meaning preservation
%   (+0.31 standardised units, p = 0.06-0.08) that approaches but does not
%   reach conventional significance; we interpret this as suggestive support
%   for H2, strengthened by convergent qualitative evidence of span-anchored
%   rewriting, rather than as a confirmed effect."
% NEVER write "significantly better" for RQ2. NEVER claim composite
% superiority (composite is tied: z 0.123 vs 0.125).
% THEME HOOK: open: "this chapter tests the thesis's central lever
% directly"; close: structure's advantage is real but narrow, on the
% dimension E2R cares most about preserving, visible only to human eyes.
% ============================================================================

\subsection{Experimental Setup}

The RQ2 controlled comparison pairs two task variants on identical inputs and identical model:
\begin{itemize}
    \item \textbf{Agentic 3-step pipeline} (this thesis's deliverable): three sequential LLM calls that (1)~detect each figurative expression verbatim, (2)~explain its meaning in context, and (3)~transform the sentence with all figurative parts replaced, each step with structured output and inspectable intermediate state.
    \item \textbf{Monolithic single-prompt} (the control condition): a stub prompt asking only for a literal rewrite of the input sentence, with free-text output. No detection step, no explicit reasoning step, no structured output schema.
\end{itemize}

Both variants run on Llama-3.1-8B-Instruct, $T=0$, on the SemEval-2022 Task A/B idiom subset. The monolithic condition uses a one-paragraph stub.

\subsection{Automatic-Metric Comparison}

The agentic and monolithic conditions cannot be separated by automatic metrics, for two structural reasons, and this is why the human evaluation carries the RQ2 verdict rather than supporting it.

First, the conditions are not comparable on detection. The monolithic condition emits a free-text rewrite with no detection structure, so detection metrics (sentence and span F1) are undefined for it: there is nothing to score. Detection is in any case not where decomposition contributes, as Section~\ref{subsec:rq2-pipeline-detection} shows.

Second, the replacement side has no adequate reference. The task admits many valid literal paraphrases that share no surface form, and SemEval-2022 Task A supplies no gold paraphrase at all. The Task B paraphrase that can be matched to a source through its multiword expression exists for only 9 of the 27 pool sources, and where it exists it is a different sentence that happens to contain the same idiom rather than a literal rewrite of the source. Reference-based scores computed against it therefore measure cross-sentence lexical overlap, not replacement quality, and cannot decide between the two systems. This is consistent with the broader result that BLEU and BERTScore correlate only spuriously with human judgement of meaning preservation and simplicity in figurative-to-literal tasks~\cite{Alva-Manchego2021,Scialom2021}. The thesis was built around this limitation: the Direct Assessment survey (Section~\ref{subsec:rq3-human-eval}) is the load-bearing instrument for RQ2, not the automatic metrics.

\paragraph{Failure modes are visible to qualitative inspection.} Spot-checking the survey-pool items shows characteristic differences between the conditions. Monolithic outputs frequently produce \emph{whole-sentence rewrites} in which non-figurative phrases are also paraphrased. For the source ``\textit{Long story short, I opted for the Letsfit White Noise Machine because it has excellent reviews on Amazon, 14 sounds, a handy nightlight, and it only costs \$20}'', for instance, the monolithic system rewrote ``opted for'' $\to$ ``chose'' and ``handy nightlight'' $\to$ ``a light that is easy to use'', whereas the agentic system left non-figurative content unchanged and only substituted ``Long story short'' $\to$ ``To summarize''. The agentic structured output anchors the rewrite to the detected span; the monolithic system has no such anchor and generalises freely. Whether the rubric rewards this discipline (under \emph{meaning preservation}) or penalises it (under \emph{simplicity}, where more aggressive simplification scores higher) is the empirical question the survey resolves.

\paragraph{Hypothesis status.} Hypothesis H2 (decomposition $>$ monolithic) is therefore neither supported nor refuted by automatic metrics on this evaluation. It stands or falls on the human ratings across the three rubric dimensions, presented next.

\subsection{Human-Evaluation Results}
\label{subsec:rq2-human-eval-results}

% --- GUIDE (~1.5 pp, ~700 words): the H2 verdict. The full survey results
% (response statistics, QC, per-system scores on all three conditions) are
% reported in Section~\ref{subsec:rq3-human-eval-results}; this section
% presents the RQ2 cut: the paired agentic-vs-monolithic comparison at 8B.
% Must land:
%  (1) Table below: agentic +0.31z on meaning preservation (paired by
%      source, n=27; t p=0.058, Wilcoxon p=0.082): the mandatory hedge
%      applies. Raw means: meaning 81.6 vs 72.6 (+9.0 raw points).
%  (2) Composite tied (z 0.123 vs 0.125): no overall-superiority claim.
%      Dimension-specific trade-off: monolithic slightly ahead on
%      grammaticality (85.4 vs 83.1) and simplicity (64.1 vs 60.6), both
%      n.s.; connect the simplicity edge to the span-anchoring discipline
%      (the monolithic system rewrites more aggressively, cf.\ the Letsfit
%      example above).
%  (3) Interpretation: decomposition buys meaning fidelity at no composite
%      cost; exactly the trade E2R adaptation wants, since meaning errors
%      are catastrophic for the target population while insufficient
%      simplification is merely unhelpful.

\begin{table}[h!]
\centering
\small
\begin{tabular}{lrrrrr}
\hline
\textbf{Dimension} & \textbf{Agentic} & \textbf{Monolithic} & \textbf{$\Delta z$} & \textbf{$t$ $p$} & \textbf{Wilc. $p$} \\
\hline
Grammaticality        & 83.1 & 85.4 & $-0.086$ & 0.450 & 0.509 \\
Meaning preservation  & 81.6 & 72.6 & $+0.309$ & 0.058 & 0.082 \\
Simplicity            & 60.6 & 64.1 & $-0.091$ & 0.611 & 0.485 \\
\hline
Composite             &      &      & $+0.044$ & 0.722 & 0.768 \\
\hline
\end{tabular}
\caption{Agentic vs.\ monolithic (both Llama-3.1-8B) on the human-evaluation survey. The Agentic and Monolithic columns are raw 0--100 means per dimension; $\Delta z$ is the paired difference in per-annotator-standardised scores (agentic $-$ monolithic, paired by source sentence, $n=27$ sources), tested with the paired $t$-test and the Wilcoxon (Wilc.) signed-rank test. Composite is the mean of the three standardised dimensions (agentic $z=0.123$, monolithic $z=0.125$); raw means are not defined for it.}
\label{tab:rq2-human-eval}
\end{table}

The three open-weights conditions are reported in full in Section~\ref{subsec:rq3-human-eval-results}; this section isolates the RQ2 cut, the paired comparison of the agentic pipeline against the monolithic single-prompt at 8B (Table~\ref{tab:rq2-human-eval}). On meaning preservation the agentic system is rated higher, 81.6 against 72.6 on the raw 0--100 scale, a nine-point gap that remains after per-annotator standardisation and pairing by source ($n = 27$). This is the chapter's central result, and it is stated with the care its statistics require: the agentic decomposition shows a consistent directional advantage on meaning preservation (+0.31 standardised units, p = 0.06-0.08) that approaches but does not reach conventional significance (the usual $p < 0.05$ threshold); we interpret this as suggestive support for H2, strengthened by convergent qualitative evidence of span-anchored rewriting, rather than as a confirmed effect.

\paragraph{No overall-superiority claim.} The advantage does not extend to the composite. The mean of the three standardised dimensions is tied, $z = 0.123$ for the agentic system against $0.125$ for the monolithic one, a difference that is negligible and far from significant (Wilcoxon $p = 0.77$). On the other two dimensions the monolithic condition is marginally ahead, on grammaticality (85.4 against 83.1) and on simplicity (64.1 against 60.6), though neither difference approaches significance. The simplicity gap is the more interpretable of the two and points back to the span-anchoring failure mode shown earlier in this chapter, where on the ``Long story short'' example the monolithic system rewrote several non-figurative phrases along with the idiom: its free whole-sentence paraphrase reads as simpler even as it drifts from the source, while the agentic system, anchored to the detected span, changes less and so preserves meaning more faithfully while reading as marginally less simple. The two conditions trade meaning fidelity against aggressive simplification, in opposite directions.

\paragraph{The trade favours decomposition for Easy-to-Read adaptation.} The two error types do not harm the target readership equally. A replacement that quietly alters the meaning of a sentence is a serious failure for a reader who depends on the adaptation, whereas a replacement that is faithful but only moderately simpler is merely less helpful. A system that secures meaning fidelity at the cost of a few points of perceived simplicity is therefore better aligned with the goal than one that simplifies more aggressively at the risk of meaning. This is what the structural lever buys: an advantage that is real but narrow, confined to the single dimension Easy-to-Read adaptation can least afford to lose, and visible only to human raters because no automatic metric in this study could separate the conditions. H2 is supported in the qualified, dimension-specific sense the data warrant, not as a blanket claim of superiority, and the advantage sits on the replacement side of the task, not in detection, as the next section shows.

\subsection{Effect of Decomposition on Detection}
\label{subsec:rq2-pipeline-detection}

% --- GUIDE (~0.75 pp, ~300 words): the honest concession, promoted from the
% Discussion to here so it cannot be read as buried. The prose below is
% moved from the Discussion's model-dependence finding (3) and may be
% edited freely; the Discussion now keeps only a summary.

On the full evaluation sets, detection F1 is near-identical under the standalone detection-only prompt and the 3-step pipeline across both scales and both phenomena: VU metaphor 0.254/0.254 at 8B and 0.302/0.302 at 70B-AWQ; SemEval idiom 0.333/0.332 at 8B and 0.529/0.529 at 70B. All seven token-, span-, and sentence-level detection metrics match across runs to within rounding; the only residual difference is on SemEval 8B (sentence F1 0.333 vs.\ 0.332), traceable to a slightly different set of structured-output parse failures rather than to any difference in detection behaviour. The match is exact by construction up to those failures rather than coincidental: the pipeline's detection step issues the same standalone detection prompt at temperature $0$, so on every successfully-parsed example its greedy decoding returns byte-identical output and therefore identical labels. The value of the comparison is that it rules out any hidden divergence (sample shift, prompt drift) and confirms that the pipeline's detection step gets essentially the F1 the detection prompt would have gotten alone, regardless of scale: whatever the pipeline architecture buys, it is not detection sharpness. This also attributes the single-call CoT's detection lift over both of these conditions to the structured-output schema's commitment scaffold rather than to chain-of-thought reasoning content per se, and it implies that decomposition's value, where it exists, must be sought on the replacement side: which is precisely where the human evaluation (Section~\ref{subsec:rq2-human-eval-results}) locates it.

\subsection{Further Comparisons}

The current RQ2 evaluation contrasts only two conditions (full agentic vs.\ monolithic). Two further comparisons are planned and discussed as future work in Chapter~\ref{ch:discussion}:
\begin{itemize}
    \item \textbf{No-explanation step}: agentic structured output without the chain-of-thought reasoning field, isolating the role of explicit reasoning;
    \item \textbf{No-self-verification}: agentic pipeline without a final consistency check between the detected span and the produced replacement.
\end{itemize}
These comparisons require minor schema changes on the existing LangGraph workflow and are deferred to give the human-evaluation survey priority.

\subsection{Failure Traceability}

A secondary benefit of the agentic decomposition is that each error mode is attributable to a specific intermediate step. The structured output records the detected expression, the explanation of its meaning, and the produced replacement; failures localise to one of these. In the monolithic condition, by contrast, errors are ``the rewrite is wrong'' with no further structure. This argument is developed in the observability discussion of Chapter~\ref{ch:discussion}.
